---
title: "Malaria-LandUse-Model"
author: "Andres Baeza"
date: "Oct 18, 2016"
output: html_document
---

#Introduction
In this document we describe how to run a single simulation of the malaria land-use model described in _Baeza etal._(_in review_) and the analyses for multiple simulations. The model considers the interactive dynamics of malaria and the human economy in a region that begins as an undeveloped state ($A_{U}$).  Development proceeds as the land is transformed at a daily rate ($l$) into a developed state ($A_{D}$) (figure 1). Simultaneous to the transformation of the land, an epidemiological human-mosquito-malaria model and a macroeconomic model that mimics a labor-intensive productive system are also at play, acting in a dynamic and interconnected way. The model is written as a system of differential equations with three modules per region that recreate the processes described in the main text and supporting information. Figure 1 shows a diagram of the epidemiological model in each land-type $U$ and $D$. The second part of the code provide the functions to analyze the model using same data as in the results of the main text, and generate new scenarios.   



####Figure 1: The epidemiological model for the two land use types
In each region the human population is divided into epidemiological classes. In untransformed or undeveloped area, the population is divided in 3 classes: susceptible ($S_{U}$) infected ($I_{U}$) and Recovered ($R_{U}$) individuals. Transformed or developed areas can include a class of individuals that do not contract malaria due to better socio-economic conditions. We called them the economically protected class ($P_{D}$). The rate of economic protection is proportional to the capital accumulation $C$. Migration increases population density and susceptible individuals. Economic as well as ecological processes can occurred at different rate in transformed and untransformed regions.

```{r,echo=TRUE,fig.align='center'}
require(png)     #read png files
require(grid)    #graphics layout
img <- readPNG("C:/Users/abaezaca/Documents/Malaria_Economy/Figure_S1_epiModel.png")
grid.raster(img)
```

#1) To read a single simulation
1.1) Read source files with differential equation models
```{r}
require(deSolve) # load Solve differential equations
source('~/Malaria_Economy/odes_malaria_LUCH.R') # Source code of model without land transformation (model_1)
source('~/Malaria_Economy/odes_malaria_LUCH_l.R') # Source code of model with land transformation (model_2)

#Note: replace "~/Malaria_Economy/" to reflect the path where the files "odes_malaria_LUCH.R" and "odes_malaria_LUCH_l.R" were saved. (e.g. source('~/PATH-To/YOUR-STUFF/odes_malaria_LUCH.R'))
```




1.2) Define parameter's values to run a single simulation
```{r}
parms_LUch<- c(
        phi=6.0,                        #Total Factor of Productivity
        alpha=0.18,                     #Elasticity capital
        beta=0.22,                      #Elasticity land
        price=4.2,                      #Average crops price
        costo_p_area=1.7,               #Cost per area planted     
        cost_p_case=2.6,                #Cost per case of malaria
        saving_rate=0.003,              #Daily income saving rate 
        depreciation=0.002,             #Daily capital depreciation rate
        fecundity_M=115,                #Mosquito fecundity
        carrying_capacity_M=0.002,      #Mosquito carrying capacity [mosquitoes /mt2]
        HBI=0.368,                      #Human biting Index [% blood meals from humans]
        c=0.5,                          #Probability infection from human to mosquito
        b=0.12,                         #Probability infection from mosquito to human
        longevity=27,                   #Natural mosquito mortality rate
        growth_Rate_H= 1 / (55 * 365),  #Human population growth rate
        mort_Rate_H=1/(55 * 365),       #Natural human mortality rate
        natutal_rec.rate=500,           #average number of days at which infected individuals acquire immunity [days]
        treatment_rec.rate=0.07,        #Rate at which infected individuals go back to susceptible per dollar invested in treatment 
        goto_Eprotect=0.05,             #Rate at which people gain protection against malaria due to socio-economic condition impruvement 
        goto_poor=0.008,                #Rate people lost protection again disease due to poverty
        delta_Env=3,                    #Degree or magnitude of ecological differences between under- and developed regions
        delta_Tech=4,                   #Degree or magnitude of productivity differences between under- and developed regions
        movement=0.006,                 #Proportion of daily movement of people between regions [% ]
        sigma=1/(7 *365),               #Loss of immunity [1/time in recovery class before becoming susceptible again [years]
        ww=0.33,                        #Debilitating effect of disease on the economy [%]
        density_U=0.0002,               #Population per unit of area in undeveloped region [people/m2]
        density_D=0.007,                #Population per unit of area in developed region [people/m2]
        l=2.026865e-05,                 #Rate of land transformation, from AU to AD  [m2/day]
        v=0.64,                         #Proportion of migrants susceptible to malaria [%]
        g=1.25                          #Gonotrophyc cycle [days]
)
```


1.3) Define State variables and initial conditions
```{r}
      #INITIAL CONDITIONS
      A_T<- 3.523e+9                                             #Total Area [km2]*[mts2/km]= 3523 * 1e6
      AU_p<-1                                                    #Proportion of area untransformed    
      AU=round(AU_p*A_T,0)                                       #Total untransformed area in hectares
      AD<- round((1-AU_p)*A_T,0)                                 #Total developed area in hectares
    
      I0_p=0.1                                                   #Initial infected population in untransformed land U
      I0_d_p=0                                                   #Initial infected population in developed land D

      S0<-AU*parms_LUch[["density_U"]]*(1-I0_p)                  #Susceptible population U
      I0<-AU*parms_LUch[["density_U"]]*I0_p                      #Infected population U
      R0<-0                                                      #Recovered (immune) population U
      P0<-0                                                      #Economically protected individuals U (in the untransformed region P=0)
      

      S0_D<-AD*parms_LUch[["density_D"]]*(1-I0_d_p)              #Susceptible population in Developed land
      I0_D<-AD*parms_LUch[["density_D"]]*I0_d_p                  #Infected population in D
      R0_D<-AD*parms_LUch[["density_D"]]*0.0                     #Recovered population in D
      P0_D<-AD*parms_LUch[["density_D"]]*0.0                     #Economically protected  population in D
      
      M0<-AU*(parms_LUch[["carrying_capacity_M"]] /3 )           #Mosquito population in untransformed land (pop. is at a third of it maximal carrying capacity)
      M0_D<-0                                                    #Mosquito pop. in developed area (M=0)
      
      X0<- M0*(1-I0_p)                                           #Population of "healthy" mosquitos (equal to prop. infected humans)
      W0<-M0*I0_p                                                #Population of infected mosquitos (equal to infected humans)
      
      X0_D<-0                                                    #Population of healthy mosquitos in D (M=0)
      W0_D<-0                                                    #Population of infected mosquitos in D (M=0)
      
      
      C0<-1*AU*parms_LUch[["density_U"]]                         #Initial capital in undeveloped land U
      C0_D<-10*parms_LUch[["l"]]*AU*parms_LUch[["density_D"]]    #Initial capital at the beginning of the land transformation  
      

      x0_NO_LUch=c(
                   C = C0,                                    
                   X = X0,
                   W = W0, 
                   S = S0, 
                   I = I0,                                        # set initial condition of model without land use change 
                   R = R0,
                   P = P0,
                   A = AU
                   )
```


1.4) Run the model without changes in land-use for 100 years
```{r}
#run model without land-use change
x_NO_LUch <- lsoda(
  y=x0_NO_LUch,
  times=seq(from=1,to=(100*365),by=1), #time in days
  func=model_1,
  parms=parms_LUch,
  hmax=0.1
)
```

1.5)Set initial conditions to run model with land-use change
```{r}
x0_LUch=c(
          C = mean(tail(x_NO_LUch[,'C'])), 
          X = mean(tail(x_NO_LUch[,'X'])),
          W = mean(tail(x_NO_LUch[,'W'])), 
          S = mean(tail(x_NO_LUch[,'S'])), #The final state of the simulation without LU change is used as initial condition for the model with transformation  
          I = mean(tail(x_NO_LUch[,'I'])), 
          R = mean(tail(x_NO_LUch[,'R'])),
          P = mean(tail(x_NO_LUch[,'P'])),
          C_d = C0_D,
          X_d = X0_D,
          W_d = W0_D,
          S_d = S0_D,
          I_d = I0_D,
          R_d = R0_D,
          P_d = P0_D,
          A = AU,
          A_d = AD
          )
```

1.6)Run model with land use change for 100 years
```{r}
##run the model with land-use change using function lsoda
  x_LUch <- lsoda(
  y=x0_LUch,
  times=seq(from=1,to=(100*365),by=1),
  func=model_2,
  parms=parms_LUch,
  hmax=0.1
)
```

1.7)Plot incidence pre and post transformation
```{r,fig.align='center'}
plot(ts(c(tail(x_NO_LUch[,'I']/rowSums(x_NO_LUch[,c('I','S','R')]),n =  (10*365)),head(rowSums(x_LUch[,c('I','I_d')])/rowSums(x_LUch[,c('I','S','R','P','I_d','S_d','R_d','P_d')]),n = (30*365))),start = 0,end = 39,frequency = 365),ylab = "Incidence [I/N]",main="Incidences of malaria pre- and post- land use change")

abline(v = 10,col="magenta") #when the tranformatinal process begins

```  
   
   
#2) Analyses of the model   

2.6) Read the outputs presented in the main text using:
To replicate results presented in the main text read the file "Results_Revision":
```{r}   
load("~/Malaria_Economy/Results_Revision") #load outputs obtained using from parameters explored in the main document
```


2.7)To select the set of simulations resulting in $I^{*}_{pre} > 0$ simply run:
```{r}   
Results_Parameters_Revision<-subset(Results_Parameters_Revision,Initial_I > 0.01 & W_mosquitoes > 0.01) 
```


2.8) Run a sensitivity analysis using the differences $I^{*}_{diff}=I^{*}_{post} - I^{*}_{pre}$ and $I^{*}_{peak-diff}=I^{*}_{peak} - I^{*}_{pre}$ (as in the main text):
```{r,fig.height=6,fig.width=16,fig.align='center'}   
require(sensitivity)   #load package sensitivity

par(mfrow=c(1,2),cex.main=2.0,cex.axis=2.0,cex.label =2) #set the figure stage for two plots

x<-pcc(X=Results_Parameters_Revision[,-(1:11)],y=(Results_Parameters_Revision$final_I -Results_Parameters_Revision$Initial_I),rank=F)    #calculate patial correlation
barplot(as.vector(x$PCC[[1]]),names.arg = names(Results_Parameters_Revision[,-(1:11)]),cex.names=0.7,las=2,main='Long term differences (I*_Diff)',ylab=expression(rho))   # barplot 1


convex_dy=subset(Results_Parameters_Revision, type_dynamics == 3 | type_dynamics == 5) #select simulations defined as "convex"
x<-pcc(X=convex_dy[,-(1:11)],y=convex_dy$max_I - convex_dy$Initial_I,rank=F)  #calculate patial correlation
barplot(as.vector(x$PCC[[1]]),names.arg = names(convex_dy[,-(1:11)]),cex.names=0.7,las=2,main='Transient differences (I*_peak)',ylab=expression(rho))  # barplot 2
```     



2.1) To generate a different set of parameter values, in the same range as in the main text.  
```{r}   
require(tgp) #load package tgp
s2_new <- lhs(10000,rbind(                                               #generate the limits of the hypercube
  c(1,10),                #Total Factor of Productivity phi
  c(0.1,0.3),             #Elasticity capital alpha
  c(0.1,0.3),             #Elasticity land beta
  c(1,10),                #Average crops price price
  c(1,5),                 #Cost per area planted costo_p_area
  c(1,6),                 #Cost per case of malaria cost_p_case
  c(0.001,0.01),          #Daily income saving rate  saving_rate
  c(0.001,0.004),         #Daily capital depreciation rate depreciation 
  c(50,200),              #Mosquito fecundity fecundity_M
  c(0.001,0.01),          #Mosquito carrying capacity [M/m2] carrying_capacity_M,
  c(0.2,1),               #Human biting/blood Index  HBI
  c(0.01,0.6),            #Probability infection from human to mosquito cc
  c(0.01,0.6),            #Probability infection from mosquito to human b
  c(12.5,30),             #Natural mosquito mortality rate longevity
  c(30,(3*365)),          #Average time  a person is infectious 
  c(0,0.1),               #Rate at which infected individuals go back to susceptible per dollar invested in treatment 
  c(0.001,0.1),           #Rate at which people gain protection against malaria due to improvement in socio-economic conditionss 
  c(0.0001,0.01),         #Rate at which infected individuals go back to susceptible do to degradation of economic investment. 
  c(1,4),                 #Degree or magnitude of ecological differences between under- and developed regions
  c(1,5),                 #Degree or magnitude of productivity differences between under- and developed regions
  c(0,0.01),              #Proportion of daily movement of people between regions
  c(1/(1*365),1/(7*365)), #Loss of immunity [1/time] in recovery class before becoming susceptible again [1/3 years]
  c(0.3,1),               #Debilitating effect of disease on the economy
  c(0.0001,0.001),        #Population per unit of area in undeveloped region  
  c(0.001,0.01),          #Population per unit of area in developed region  
  c(0,1),                 #Proportion of migrants susceptible to malaria  
  c(1,4),                 #Gonotrophyc cycle
  c(1.5e-05,3e-05)        #Land use change rate
)
)
```   

2.2) Add the names of the outputs to save in matrix's columns
```{r} 
colnames(s2_new)<-c(
  "phi",
  "alpha",
  "beta",
  "price",
  "costo_p_area",
  "cost_p_case",
  "saving_rate",
  "depreciation",
  "fecundity_M", 
  "carrying_capasity_M",
  "HBI",
  "prob_inf_HM",
  "prob_inf_MH",
  "mosquito_lifetime",
  "days_person_infectious",
  "treatment",
  "goto_reach",
  "goto_poor",
  "delta_Eco",
  "delta_Tech",
  "movement",
  "sigma",
  "w",
  "density_U",
  "density_D",
  "v",
  "g",
  "l")
```     

To explore the entire parameters' space, the model must be run in a loop for the 10000 different scenarios created before. In each cycle of the loop we calculated 1) the vector for the parameters that would run the differential equation models (step 1.2). Define the initial conditions (step 1.3). Run of the model in conditions pre-transformation (step 1.4). Set initial conditions for the model under land use change (step 1.5). Run the model under land use change (step 1.6).


2.3) To save results in each run of the model first create a table with the dimensions of the hypercube, in this case (10000x11)
```{r}   
table_results=matrix(ncol=11,nrow=10000)        #define table
colnames(table_results)<-c("which.max","which.min","dif_Inc","type_dynamics","Initial_I","final_I","max_I","Capital","Capital_D","W_mosquitoes","W_mosquitoes_D")  # define names of quantities to save
```

2.4) Fill each row of table_results using model outputs from section 1 of this document
```{r}   
#generate an output vector of size 11
output=c( 
   which.max(rowSums(x_LUch[,c('I','I_d')])/rowSums(x_LUch[,c('I','S','R','P','I_d','S_d','R_d','P_d')])) #which.max
   
  ,which.min(rowSums(x_LUch[,c('I','I_d')])/rowSums(x_LUch[,c('I','S','R','P','I_d','S_d','R_d','P_d')]))#which.min
  
  ,mean(tail(x_NO_LUch[,'I']/rowSums(x_NO_LUch[,c('I','S','R','P')]),10)) - mean(tail(rowSums(x_LUch[,c('I','I_d')])/rowSums(x_LUch[,c('I','S','R','P','I_d','S_d','R_d','P_d')]),10))##I_dif
  
  ,0                  #type_dynamics defined next 
  ,mean(tail(x_NO_LUch[,'I']/rowSums(x_NO_LUch[,c('I','S','R','P')]),10))# Initial_I
  
  ,mean(tail(rowSums(x_LUch[,c('I','I_d')])/rowSums(x_LUch[,c('I','S','R','P','I_d','S_d','R_d','P_d')]),100))#final_I
  
  ,max(rowSums(x_LUch[,c('I','I_d')])/rowSums(x_LUch[,c('I','S','R','P','I_d','S_d','R_d','P_d')]))[1] #max_I
  
  ,mean(tail(x_NO_LUch[,'C']/rowSums(x_NO_LUch[,c('I','S','R','P')]),10)) #Capital
  
  ,mean(tail(rowSums(x_LUch[,c('C','C_d')])/rowSums(x_LUch[,c('I','S','R','P','I_d','S_d','R_d','P_d')]),100))#Capital_D
  
  ,mean(tail(x_NO_LUch[,'W']/rowSums(x_NO_LUch[,c('X','W')]),10))#W_mosquitoes 
  
  ,mean(tail(rowSums(x_LUch[,c('W','W_d')])/rowSums(x_LUch[,c('X','W','X_d','W_d')]),100))#W_mosquitoes_D
  
    )
print(output)
```

2.5) To Classify trajectories
fill-out column 4 ("type_dynamics") of table-results-Revision with the type of trajetory obtained (here using data presented in the main text).

```{r,echo=TRUE}   
print(output[1]==1 & output[2]>36000 & output[3]>0)                     #1: Monotonic decrease
print(output[1]>36000 & output[2]==1 & output[3]<0)                     #2: Monotonic increase
print(output[1]>1 & output[1]<36000 & output[2]>36000 & output[3]>0)    #3: Convex decrease
print(output[1]>36000 & output[2]>1 & output[2]<36000 & output[3]<0)    #4: Concave increase
print(output[1]>1 & output[1]<36000 & output[1]==1 & output[3]<0)       #5: Convex increase
print(output[1]==1 & output[2]>1 & output[2]<36000 & output[3]>0)       #6: Concave decrease
```     

2.6) Pie plots with the proportion of simulations clasified in each trajectory (as in the main text)
```{r,echo=TRUE}   
### calcula proportions###
#1) 6 types
slices_LI<-c(length(which(Results_Parameters_Revision[,4]==1)),
             length(which(Results_Parameters_Revision[,4]==2)),
             length(which(Results_Parameters_Revision[,4]==3)),
             length(which(Results_Parameters_Revision[,4]==4)),
             length(which(Results_Parameters_Revision[,4]==5)),
             length(which(Results_Parameters_Revision[,4]==6)))
    
    
slices_Dec_or_Inc<-c(length(which(Results_Parameters_Revision[,3]>0)),
                     length(which(Results_Parameters_Revision[,3]<0)))
    
#2) 3 types "convex" "concave" and "monotonic"
slices_LI3ty<-c(length(which(Results_Parameters_Revision[,4]==1))+
                length(which(Results_Parameters_Revision[,4]==2)),
                length(which(Results_Parameters_Revision[,4]==3))+length(which(Results_Parameters_Revision[,4]==5)),
               length(which(Results_Parameters_Revision[,4]==4))+                    
               length(which(Results_Parameters_Revision[,4]==6)))
  
#make the plots
names<-c('Monotonic D.','Monotonic I.',"Convex D.","Concave I.", "Convex I.","Concave D.")
par(mfrow=c(1,3))
pie(slices_Dec_or_Inc,labels=c("Decrease","Increase"),cex.lab=2.0)
pie(slices_LI3ty, labels = c("Monotonic","Convex","Concave"), main="")  
pie(slices_LI, labels = names, main="")
 
   
```
